### 2018-03-09 submission/manuscript.tex
change : to . for titles

\textbf{\emph{Lower bacterial diversity is associated with increased
odds ratio (OR) of tumors:}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was not a significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
diversity were only significantly greater than 1.0 for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
1.0, it is doubtful that these are clinically meaningful values.

@@

\textbf{\emph{Disease progression is associated with community-wide
changes in composition and abundance:}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there consistent and significant community-wide changes in
the fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual taxa are associated with significant ORs for
carcinomas:}} Next we identified those taxa were associated with ORs
that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
0.83)). Among the data collected from tissue samples, only unmatched
carcinoma samples had taxa with a significant OR. Those included
\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
Mouth-associated genera were not significantly associated with an
increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
due to the small number of studies and considerable variation in the
data, the Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}.
It is interesting to note that \emph{Ruminococcus} and members of
\emph{Clostridium group XI} in stool and \emph{Dorea} and \emph{Blautia}
in tissue had ORs that were significantly less than 1.0, which suggests
that these populations are protective against the development of
carcinomas. Overall, there was no overlap in the taxa with significant
OR between stool and tissue samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma:}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. OR values were
caluclated based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study. To measure the ability of these taxa to
classify individuals we instead generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included \emph{Clostridium XI},
\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
AUC values less than 0.588. Likewise, in unmatched tissue samples the 8
taxa with significant ORs taxa were marginally better than one would
expect by chance {[}Figure 3B{]}. The relative abundance of Dorea was
the best predictor of carcinomas and its median AUC was only 0.62. These
results suggest that although these taxa are associated with a decreased
or increased OR for the presences of carcinomas, individually, they do a
poor job of classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa:}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). For the full
taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
were the most common taxa in the top 10\% mean decrease in accuracy
(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
ORs, all 3 were among the top 10\% most important taxa as measured by
mean decrease in accuracy, in at least one study. The most important
taxa across study within the significant OR taxa only models for stool
were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure 5A{]}.
Similarly, using data from the unmatched tissue samples, the combined
model had an AUC of 0.77, which was significantly higher than the AUC
values for Blautia and Weissella (P-value \textless{} 0.037). For the
full taxa models using unmatched tissue, \emph{Lachnospiraceae},
\emph{Bacteroidaceae}, and \emph{Ruminococcaceae} were the most common
taxa in the top 10\% mean decrease in accuracy across studies {[}Figure
S3{]}. For the singificant OR taxa unmatched tissue models both
\emph{Dorea} and \emph{Blautia} were the important based on mean
decrease in accuracy {[}Figure 5B{]}. Pooling the information from the
taxa with significant ORs results in a model that outperforms
classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs:}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC =
0.5486298 {[}0.3671667 - 0.971{]}) {[}Figure 4A \& S4A{]}. In contrast,
the models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of -14.8\%) and AUC for the Flemer study decreased by
0.54\%). The overall improvement in performance was statistically
significant (mean = 12.61\%, one-tailed paired T-test; P-value = 0.005).
Similarly, using the data from unmatched tissue samples we found that
the AUC of studies increased by an average of 19.11\% when we used all
of the taxa rather than the 3 taxa with significant ORs (one-tailed
paired T-test; P-value = 0.03). Although the significant OR taxa models
can classify indidivuals with and without carcinoma tumors, they are
still missing taxa from the full community models that can increase the
model accuracy.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs:}} The previous models were based on relative
abundance data where sequences were classified to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 {[}0.37- 0.56{]}), but
did a good job of differentiating between subjects with normal colons
and those with carcinomas (median AUC: 0.71 {[}0.5- 0.9{]}). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value = 0.966) and
those using all taxa (one-tailed paired T-test; P-value = 0.146)
{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
accuracy for the OTU-based models, we found that OTUs that affiliated
with all of the 8 taxa that had a significant OR were within the top
10\% for at least one study. This result was surprising as it indicated
that a finer scale classification of the sequences and thus a larger
number of features to select from, did not yield improved classification
of the subjects.

\textbf{\emph{Generalizability of taxon-based models trained on one
dataset to the other datasets:}} Considering the good performance of the
Random Forest models using taxa with a significant OR and using all of
the taxa, we next asked how well the models would perform when given
data from a different subject cohort. For instance, if a model was
trained using data from the Ahn study, we wanted to know how well it
would perform using the data from the Baxter study. We found the models
trained using the taxa with a significant OR all had a higher median AUC
than the models trained using all of the taxa when tested on the other
datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
between the performance of the modelling approaches appeared to vary
with the size of the training cohort (R\textsuperscript{2} = 0.66)
{[}Figure 6{]}. These data suggest that given a sufficient number of
subjects with normal colons and carcinomas, Random Forest models trained
using a small number of taxa can accurately classify individuals from a
different cohort.

@@

\textbf{\emph{Datasets:}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (41) and
Vogtmann, et al. (42). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
excluded because they did not have publicly available sequences or did
not have metadata in which the authors were able to share. We were able
to obtain sequence data and metadata from the following studies: Ahn,
\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
(44) study was excluded because their 90 individuals were contained
within the larger Baxter study (12). The Kostic study was excluded
because after we processed the sequences, all of the case samples had
100 or fewer sequences. The final analysis included 14 studies (Tables 1
and 2). There were seven studies with only fecal samples (Ahn, Baxter,
Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
fecal and tissue samples (Chen and Flemer). After curating the
sequences, 1737 stool samples and 492 tissue samples remained in the
analysis {[}Tables 1 and 2{]}.

\textbf{\emph{Sequence Processing:}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n = 4,
(17, 23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) (45) using the default quality filtering methods for both 454
and Illumina sequence data. If it was not possible to use the defaults
because the sequences were trimmed too much, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (46). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (47)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (48)).

\textbf{\emph{Community analysis:}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (49) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (50). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis:}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. Within the training
dataset, 10-fold cross validation (5-fold cross validation for small
datasets) was used to build a model that was then evaluated on the
testing set. For the models constructed using the taxa with significant
ORs, the default mtry setting was used to train the model and this model
was tested on the other datasets in the meta-analysis. The reported AUC
values are the AUCs for the application of the model on the test sets.
For the OTU-based models, the dataset was split into training (80\% of
samples) and testing (20\%) sets and 10-fold cross validation (5-fold
cross validation for small datasets) on the training set was used to
generate the model for the testing set. The original 80/20 split and
fitting was repeated 100 times and the average AUC from these 100
repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
the importance of each taxon to the overall model was used to rank the
taxa used in each model. For all models, the default setting used was
\(\sqrt{p}\), where p is all the variables used in the respective model.
Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
mtry (51).

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (52). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (53) and then
Z-score normalized using the car (v2.1.6) package (54). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models created using the lme4 (v1.1.15) package (55) to
correct for study, repeat sampling of individuals (tissue only), and 16S
hyper-variable region used. Odds ratios (OR) were analyzed using both
the epiR (v0.9.93) and metafor (v2.0.0) packages (56, 57) by assessing
how many individuals with and without disease were above and below the
overall median value within each specific study. OR significance testing
utilized the chi-squared test. Diversity differences measured by the
Bray-Curtis index utilized the creation of distance matrix and testing
with PERMANOVA executed with the vegan (v2.4.5) package (58). Random
Forest models were built using both the caret (v6.0.78) and randomForest
(v4.6.12) packages (59, 60). All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages (61, 62).

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number of raw sequences from
the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.

>>>>

\textbf{\emph{Lower bacterial diversity is associated with increased
odds ratio (OR) of tumors.}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was not a significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
diversity were only significantly greater than 1.0 for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
1.0, it is doubtful that these are clinically meaningful values.

@@

\textbf{\emph{Disease progression is associated with community-wide
changes in composition and abundance.}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there consistent and significant community-wide changes in
the fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual taxa are associated with significant ORs for
carcinomas.}} Next we identified those taxa were associated with ORs
that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
0.83)). Among the data collected from tissue samples, only unmatched
carcinoma samples had taxa with a significant OR. Those included
\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
Mouth-associated genera were not significantly associated with an
increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
due to the small number of studies and considerable variation in the
data, the Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}.
It is interesting to note that \emph{Ruminococcus} and members of
\emph{Clostridium group XI} in stool and \emph{Dorea} and \emph{Blautia}
in tissue had ORs that were significantly less than 1.0, which suggests
that these populations are protective against the development of
carcinomas. Overall, there was no overlap in the taxa with significant
OR between stool and tissue samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma.}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. OR values were
caluclated based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study. To measure the ability of these taxa to
classify individuals we instead generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included \emph{Clostridium XI},
\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
AUC values less than 0.588. Likewise, in unmatched tissue samples the 8
taxa with significant ORs taxa were marginally better than one would
expect by chance {[}Figure 3B{]}. The relative abundance of Dorea was
the best predictor of carcinomas and its median AUC was only 0.62. These
results suggest that although these taxa are associated with a decreased
or increased OR for the presences of carcinomas, individually, they do a
poor job of classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa.}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). For the full
taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
were the most common taxa in the top 10\% mean decrease in accuracy
(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
ORs, all 3 were among the top 10\% most important taxa as measured by
mean decrease in accuracy, in at least one study. The most important
taxa across study within the significant OR taxa only models for stool
were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure 5A{]}.
Similarly, using data from the unmatched tissue samples, the combined
model had an AUC of 0.77, which was significantly higher than the AUC
values for Blautia and Weissella (P-value \textless{} 0.037). For the
full taxa models using unmatched tissue, \emph{Lachnospiraceae},
\emph{Bacteroidaceae}, and \emph{Ruminococcaceae} were the most common
taxa in the top 10\% mean decrease in accuracy across studies {[}Figure
S3{]}. For the singificant OR taxa unmatched tissue models both
\emph{Dorea} and \emph{Blautia} were the important based on mean
decrease in accuracy {[}Figure 5B{]}. Pooling the information from the
taxa with significant ORs results in a model that outperforms
classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs.}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC =
0.5486298 {[}0.3671667 - 0.971{]}) {[}Figure 4A \& S4A{]}. In contrast,
the models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of -14.8\%) and AUC for the Flemer study decreased by
0.54\%). The overall improvement in performance was statistically
significant (mean = 12.61\%, one-tailed paired T-test; P-value = 0.005).
Similarly, using the data from unmatched tissue samples we found that
the AUC of studies increased by an average of 19.11\% when we used all
of the taxa rather than the 3 taxa with significant ORs (one-tailed
paired T-test; P-value = 0.03). Although the significant OR taxa models
can classify indidivuals with and without carcinoma tumors, they are
still missing taxa from the full community models that can increase the
model accuracy.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs.}} The previous models were based on relative
abundance data where sequences were classified to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 {[}0.37- 0.56{]}), but
did a good job of differentiating between subjects with normal colons
and those with carcinomas (median AUC: 0.71 {[}0.5- 0.9{]}). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value = 0.966) and
those using all taxa (one-tailed paired T-test; P-value = 0.146)
{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
accuracy for the OTU-based models, we found that OTUs that affiliated
with all of the 8 taxa that had a significant OR were within the top
10\% for at least one study. This result was surprising as it indicated
that a finer scale classification of the sequences and thus a larger
number of features to select from, did not yield improved classification
of the subjects.

\textbf{\emph{Generalizability of taxon-based models trained on one
dataset to the other datasets.}} Considering the good performance of the
Random Forest models using taxa with a significant R and using all of
the taxa, we next asked how well the models would perform when given
data from a different subject cohort. For instance, if a model was
trained using data from the Ahn study, we wanted to know how well it
would perform using the data from the Baxter study. We found the models
trained using the taxa with a significant OR all had a higher median AUC
than the models trained using all of the taxa when tested on the other
datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
between the performance of the modelling approaches appeared to vary
with the size of the training cohort (R\textsuperscript{2} = 0.66)
{[}Figure 6{]}. These data suggest that given a sufficient number of
subjects with normal colons and carcinomas, Random Forest models trained
using a small number of taxa can accurately classify individuals from a
different cohort.

@@

\textbf{\emph{Datasets.}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (41) and
Vogtmann, et al. (42). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
excluded because they did not have publicly available sequences or did
not have metadata in which the authors were able to share. We were able
to obtain sequence data and metadata from the following studies: Ahn,
\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
(44) study was excluded because their 90 individuals were contained
within the larger Baxter study (12). The Kostic study was excluded
because after we processed the sequences, all of the case samples had
100 or fewer sequences. The final analysis included 14 studies (Tables 1
and 2). There were seven studies with only fecal samples (Ahn, Baxter,
Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
fecal and tissue samples (Chen and Flemer). After curating the
sequences, 1737 stool samples and 492 tissue samples remained in the
analysis {[}Tables 1 and 2{]}.

\textbf{\emph{Sequence Processing.}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n = 4,
(17, 23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) (45) using the default quality filtering methods for both 454
and Illumina sequence data. If it was not possible to use the defaults
because the sequences were trimmed too much, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (46). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (47)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (48)).

\textbf{\emph{Community analysis.}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (49) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (50). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis.}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. Within the training
dataset, 10-fold cross validation (5-fold cross validation for small
datasets) was used to build a model that was then evaluated on the
testing set. For the models constructed using the taxa with significant
ORs, the default mtry setting was used to train the model and this model
was tested on the other datasets in the meta-analysis. The reported AUC
values are the AUCs for the application of the model on the test sets.
For the OTU-based models, the dataset was split into training (80\% of
samples) and testing (20\%) sets and 10-fold cross validation (5-fold
cross validation for small datasets) on the training set was used to
generate the model for the testing set. The original 80/20 split and
fitting was repeated 100 times and the average AUC from these 100
repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
the importance of each taxon to the overall model was used to rank the
taxa used in each model. For all models, the default setting used was
\(\sqrt{p}\), where p is all the variables used in the respective model.
Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
mtry (51).

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (52). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (53) and then
Z-score normalized using the car (v2.1.6) package (54). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models created using the lme4 (v1.1.15) package (55) to
correct for study, repeat sampling of individuals (tissue only), and 16S
hyper-variable region used. Odds ratios (OR) were analyzed using both
the epiR (v0.9.93) and metafor (v2.0.0) packages (56, 57) by assessing
how many individuals with and without disease were above and below the
overall median value within each specific study. OR significance testing
utilized the chi-squared test. Diversity differences measured by the
Bray-Curtis index utilized the creation of distance matrix and testing
with PERMANOVA executed with the vegan (v2.4.5) package (58). Random
Forest models were built using both the caret (v6.0.78) and randomForest
(v4.6.12) packages (59, 60). All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages (61, 62).

\textbf{\emph{Reproducible Methods.}} The code and analysis can be found
at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number of raw sequences from
the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.
